\chapter*{Acknowledgements}
\addcontentsline{toc}{chapter}{Acknowledgments}
\begin{center}
 \textit{``If I have seen a little further it is by standing on the shoulders of giants.''}
\end{center}
\begin{flushright}
 Sir Isaac Newton
\end{flushright}

This thesis can never be compared to the works of Newton, but it has one thing in common with them: It would never have been possible without the achievements and support of others. Here, I'd like to thank all the 'giants' that have let me stand on their shoulders the past six months:

First of all I'd like to thank my supervisor Andreas Leitner who was never tired of discussing with me and clarifying things. Most design decisions, I made based on his council and he ensured that I never lost the thread on my way. He also helped to get the report that you hold in your hands into the right form, by investing hours on cross-reading and giving me advice.

Emmanuel Stapf of Eiffelsoftware inc. provided many useful tips about how to implement requested features. It always seemed that ten minutes of his time are equal to two weeks of mine, undoubtedly he saved me from poking in the countless lines of code that Eiffel studio contains, speeding up the development time enormously.

I'd like to thank Prof. Dr. Bertrand Meyer for making this work possible at all, first by designing the language Eiffel, which is still a class of its own and of course by letting me work for his group.

All members of the software engineering group who never hesitated to help me out, whenever I was stuck.

My co students Arno Fiva and Marco Zietzling: Arno for working together with me on the integration of selective capture and replay into Cdd and Marco for providing me his \LaTeX template for this report.

Alex Orso, one of the authors of the original paper about selective capture and replay \cite{orso05may} for quickly answering my questions regarding the technique.

Finally I want thank my family for supporting me throughout all my years of study and my girlfriend for cheering me up during this thesis; it is them who made all this possible.